Viscous Transport in Eroding Porous Media

### 2019-09-27 docs/transport/erosion_transport.tex
small edits from S-H

We present the quadrature rules for a simply-connected interior domain
$\Omega \subset \CC$, with a point $a \in \Omega$, and we consider the
target points $x \in \Omega$ and $x \in \Omega^c$. Then, these
quadrature rules can be applied to individual components of a
multiply-connected domain to compute layer potentials of a porous media.
The method starts with an underlying quadrature rule, and we use the
spectrally accurate trapezoid rule.  Since the quadrature points are
uniformly distributed, the quadrature weights are $w_j = L/N$,
$j=1,\ldots,N$, where $L$ is the length of $\bd\Omega$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Efficiently Applying the Quadrature}
\label{sec:fmm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By using the Barycentric quadrature rule, the velocity, shear stress,
and vorticity are computed with an error that is bounded independent of
the target location. However, applied directly, it requires
$\mathcal{O}(N^2)$ operations.  By using a fast summation method, such
as the fast multipole method~\cite{gre-rok1987}, this cost can be
reduced to $\mathcal{O}(N)$ operations.  However, the Barycentric
quadrature rules require several $N$-body calculations, and they must be
done for each component of $\bd\Omega$ individually.  The resulting cost
would be prohibitive, so we introduce a hybrid method that combines the
Barycentric quadrature rule and the trapezoid rule.  Note that the
source points of the layer potential is always one of the eroding bodies
or the outer wall, but the target point can either be on another
component of $\bd\Omega$ or it can be in the fluid bulk $\Omega$.

To compute the velocity double-layer potential, we start by applying the
trapezoid rule~\eqref{eqn:trap} accelerated with the FMM.  This
calculation requires $\mathcal{O}(N)$ operations, and we call the
resulting velocity $\vv_\trap(\xx)$.  Since the trapezoid rule is
spectrally accurate, the error of $\vv_\trap(\xx)$ is small for points
sufficiently far from $\bd\Omega$, and this region depends on the number
of discretization points $N_\iin$ and $N_\out$.  However, the trapezoid
rule needs to be replaced with a more accurate quadrature rule for
points that are too close to $\bd\Omega$.  Note, however, that since a
point is typically only close to one or two components of $\bd\Omega$,
only the contribution of these nearby bodies needs to be replaced.
Assuming that $\xx$ is too close to $\gamma_k$, we first subtract the
inaccurate trapezoid rule approximation of the double-layer potential
due to $\gamma_k$.  Then, the Barycentric quadrature rule is used to
compute the velocity due to $\gamma_k$ with more accuracy.  That is, the
final velocity is
\begin{align}
  \vv(\xx) = \vv_\trap(\xx) - \frac{L_k}{N_k} \sum_{j=1}^{N_\iin} 
    D(\xx,\yy^k_j) \eeta(\yy_j) + \vv^k_\bary(\xx),
  \label{eqn:velocityDecomp}
\end{align}
where $\vv^k_\bary(\xx)$ is the velocity at $\xx$ resulting from
applying the Barycentric quadrature rule to the double-layer potential
due to $\gamma_k$.  This strategy naturally extends to points that are
close to $\Gamma$, and to the scenario when $\xx$ is simultaneously
close to multiple components of $\bd\Omega$.  While the term
$\vv_\trap(\xx)$ in equation~\eqref{eqn:velocityDecomp} is computed for
all target points using the FMM, the other two terms are computed with a
direct summation.  However, these terms are only required for target
points near an eroding body or the outer wall, and these points make up
only a small fraction of the total number of points.

@@

Per grid point, applying the Barycentric quadrature rules dominates the
computational cost. Therefore, it is imperative that we only apply the
Barycentric quadrature when the trapezoid rule is inaccurate.  As a rule
of thumb, the trapezoid rule due to $\gamma_k$ achieves machine epsilon
accuracy if~\cite{bar2014}
\begin{align}
  \label{eqn:TrapCutoff}
  d(\xx,\gamma_k) = \inf_{\yy \in \gamma_k} \|\xx - \yy\| > 
    5 \frac{L_k}{N_\iin}.
\end{align}
To avoid checking if all target point $\xx$
satisfy~\eqref{eqn:TrapCutoff}, we make use of the {\thL}
variables~\cite{hou-low-she1994}.  First, given two eroding bodies $\gamma_i$ and $\gamma_j$ with
lengths $L_i$ and $L_j$, and centers $\cc_i$ and $\cc_j$, we check if
\begin{align}
  \label{eqn:BodyCutoff}
  \|\cc_i - \cc_j\| < \frac{L_i}{2\pi} + \frac{L_j}{2\pi} + 
    \alpha_\iin \left(\frac{L_i}{N_\iin} + \frac{L_j}{N_\iin} \right),
\end{align}
where $\alpha_\iin \geq 1$ is a parameter that needs to be determined.
In this manner, rather than using an expensive all-to-all algorithm to
compute the distance between all pairs of discretization points, we
compute the distance between pairs of circle centers.  This criteria
allows us to quickly determine bodies that contain discretization points
where the Barycentric quadrature rule might need to be applied, and the
parameter $\alpha_\iin$ accounts for the fact that the bodies are being
approximated with circles.  Assuming that the two bodies $\gamma_i$ and
$\gamma_j$ satisfy condition~\eqref{eqn:BodyCutoff}, for each point $\xx
\in \gamma_j$, we check if
\begin{align}
  \label{eqn:PointsBodyCutoff}
  \|\xx - \cc_i\| < \frac{L_i}{2\pi}
+ \alpha_\iin \left(\frac{L_i}{N_\iin} \right).
\end{align}
To determine if points on $\gamma_i$ are too close to the outer wall, we
recall that the eroding bodies are all contained in $[-1,1] \times
[-1,1]$, so a target point can only be close to the lines $y = \pm 1$.
Therefore, we first check if
\begin{align}
  \left\|\cc_i - \left[
    \begin{array}{c}
      x_i \\ \pm 1
    \end{array}
    \right]
  \right\| < \frac{L_i}{2 \pi} + \alpha_\out \frac{L_\out}{N_\out},
\end{align}
where $x_i$ is the $x$-coordinate of $\cc_i$. If body $\gamma_i$
satisfies this condition, for each point $\xx=(x,y) \in \gamma_i$, we
apply the Barycentric rule to points that satisfy
\begin{align}
  \label{eqn:PointsWallCutoff}
  |y \pm 1| < \frac{L_i}{2 \pi} +\alpha_\out \frac{L_\out}{N_\out}.
\end{align}
Finally, to determine if a target point $\xx$ in the fluid bulk requires
the Barycentric quadrature rule, we only check
conditions~\eqref{eqn:PointsBodyCutoff}
and~\eqref{eqn:PointsWallCutoff}.

>>>>

We present the quadrature rules for a simply-connected interior domain
$\Omega \subset \CC$, with any point $a \in \Omega$, and we consider the
target points $x \in \Omega$ and $x \in \Omega^c$. Then, these
quadrature rules can be applied to individual components of a
multiply-connected domain to compute layer potentials of a porous media.
The method starts with an underlying quadrature rule, and we use the
spectrally accurate trapezoid rule.  Since the quadrature points are
uniformly distributed, the quadrature weights are $w_j = L/N$,
$j=1,\ldots,N$, where $L$ is the length of $\bd\Omega$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Efficiently Applying the Quadrature}
\label{sec:fmm}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
By using the Barycentric quadrature rule, the velocity, shear stress,
and vorticity are computed with an error that is bounded independent of
the target location. However, applied directly, it requires
$\mathcal{O}(N^2)$ operations.  By using a fast summation method, such
as the fast multipole method (FMM)~\cite{gre-rok1987}, this cost can be
reduced to $\mathcal{O}(N)$ operations.  However, the Barycentric
quadrature rules require several $N$-body calculations, and they must be
done for each component of $\bd\Omega$ individually.  The resulting cost
would be prohibitive, so we introduce a hybrid method that combines the
Barycentric quadrature rule and the trapezoid rule.  Note that the
source points of the layer potential is always one of the eroding bodies
or the outer wall, but the target point can either be on another
component of $\bd\Omega$ or it can be in the fluid bulk $\Omega$.

To compute the velocity double-layer potential, we start by applying the
trapezoid rule~\eqref{eqn:trap} accelerated with the FMM.  This
calculation requires $\mathcal{O}(N)$ operations, and we call the
resulting velocity $\vv_\trap(\xx)$.  Since the trapezoid rule is
spectrally accurate, the error of $\vv_\trap(\xx)$ is small for points
sufficiently far from $\bd\Omega$, and this region depends on the number
of discretization points $N_\iin$ and $N_\out$.  However, the trapezoid
rule needs to be replaced with a more accurate quadrature rule for
points that are too close to $\bd\Omega$.  Note that since a point is
typically only close to one or two components of $\bd\Omega$, only the
contribution of these nearby bodies needs to be replaced.  Assuming that
$\xx$ is too close to $\gamma_k$, we first subtract the inaccurate
trapezoid rule approximation of the double-layer potential due to
$\gamma_k$.  Then, the Barycentric quadrature rule is used to compute
the velocity due to $\gamma_k$ with more accuracy.  That is, the final
velocity is
\begin{align}
  \vv(\xx) = \vv_\trap(\xx) - \frac{L_k}{N_k} \sum_{j=1}^{N_\iin} 
    D(\xx,\yy^k_j) \eeta(\yy_j) + \vv^k_\bary(\xx),
  \label{eqn:velocityDecomp}
\end{align}
where $\vv^k_\bary(\xx)$ is the velocity at $\xx$ resulting from
applying the Barycentric quadrature rule to the double-layer potential
due to $\gamma_k$.  This strategy naturally extends to points that are
close to $\Gamma$, and to the scenario when $\xx$ is simultaneously
close to multiple components of $\bd\Omega$.  While the term
$\vv_\trap(\xx)$ in equation~\eqref{eqn:velocityDecomp} is computed for
all target points using the FMM, the other two terms are computed with a
direct summation.  However, these terms are only required for target
points near an eroding body or the outer wall, and these points make up
only a small fraction of the total number of points.

@@

Per grid point, applying the Barycentric quadrature rules dominates the
computational cost. Therefore, it is imperative that we only apply the
Barycentric quadrature when the trapezoid rule is inaccurate.  As a rule
of thumb, the trapezoid rule due to $\gamma_k$ achieves machine epsilon
accuracy if~\cite{bar2014}
\begin{align}
  \label{eqn:TrapCutoff}
  d(\xx,\gamma_k) = \inf_{\yy \in \gamma_k} \|\xx - \yy\| > 
    5 \frac{L_k}{N_\iin}.
\end{align}
To avoid checking if all target point $\xx$
satisfy~\eqref{eqn:TrapCutoff}, we make use of the {\thL}
variables~\cite{hou-low-she1994}.  First, given two eroding bodies
$\gamma_i$ and $\gamma_j$ with lengths $L_i$ and $L_j$, and centroids
$\cc_i$ and $\cc_j$, we check if
\begin{align}
  \label{eqn:BodyCutoff}
  \|\cc_i - \cc_j\| < \frac{L_i}{2\pi} + \frac{L_j}{2\pi} + 
    \alpha_\iin \left(\frac{L_i}{N_\iin} + \frac{L_j}{N_\iin} \right),
\end{align}
where $\alpha_\iin \geq 1$ is a parameter that needs to be determined.
In this manner, rather than using an expensive all-to-all algorithm to
compute the distance between all pairs of discretization points, we
compute the distance between pairs of circle centers.  This criteria
allows us to quickly determine bodies that contain discretization points
where the Barycentric quadrature rule might need to be applied, and the
parameter $\alpha_\iin$ accounts for the fact that the bodies are being
approximated with circles.  Assuming that the two bodies $\gamma_i$ and
$\gamma_j$ satisfy condition~\eqref{eqn:BodyCutoff}, for each point $\xx
\in \gamma_j$, we check if
\begin{align}
  \label{eqn:PointsBodyCutoff}
  \|\xx - \cc_i\| < \frac{L_i}{2\pi}
+ \alpha_\iin \left(\frac{L_i}{N_\iin} \right).
\end{align}
To determine if points on $\gamma_i$ are too close to the outer wall, we
recall that the eroding bodies are all contained in $[-1,1] \times
[-1,1]$, so a target point can only be close to the lines $y = \pm 1$.
Therefore, we first check if
\begin{align}
  \left\|\cc_i - \left[
    \begin{array}{c}
      x_i \\ \pm 1
    \end{array}
    \right]
  \right\| < \frac{L_i}{2 \pi} + \alpha_\out \frac{L_\out}{N_\out},
\end{align}
where $x_i$ is the $x$-coordinate of $\cc_i$. If body $\gamma_i$
satisfies this condition, for each point $\xx=(x,y) \in \gamma_i$, we
apply the Barycentric rule to points that satisfy
\begin{align}
  \label{eqn:PointsWallCutoff}
  |y \pm 1| < \frac{L_i}{2 \pi} +\alpha_\out \frac{L_\out}{N_\out}.
\end{align}
Finally, to determine if a target point $\xx$ in the fluid bulk requires
the Barycentric quadrature rule, we only check
conditions~\eqref{eqn:PointsBodyCutoff}
and~\eqref{eqn:PointsWallCutoff}.
